Claret: Using Data Types for Highly Concurrent Distributed Transactions

### 2015-02-14 rmd/paper.Rmd
fix references

In situations where consistency really is crucial, developers can enforce stronger guarantees on top of any platform, but they often get it wrong. Google, recognizing the benefit to programmer productivity, reintroduced serializable transactions to their next generation database, Spanner \cite{Spanner}. However, this leaves them with two extreme choices, with significant performance impact. If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model? What if the programmer could express the semantics they desire succintly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?

@@

Performance benefits come from understanding the properties of ADT operations: those that commute with each other can be performed concurrently, even on multiple copies of the record. This means that transactions whose operations commute abort less, approaching the performance without transactions. This cleanly captures the cases described above where conflicts were unobservable or ordering didn't matter but in a safe way because any operations that don't in fact commute will be handled with traditional concurrency control. Using insights from multi-core concurrent data structures, we show in Section \ref{sec:commutativity} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.

@@

# Commutativity {#comm}
\label{sec:commutativity}

@@

\begin{table}
\centering
\begin{tabular}{lll}
\textbf{method:} & \textbf{commutes with:} & \textbf{when:} \\
\hline
\texttt{add(x): void} & \texttt{add(y)} & $\forall x, y$ \\
\texttt{remove(x): void} & \texttt{remove(y)} & $\forall x, y$ \\
    & \texttt{add(y)} & $x \ne y$ \\
\texttt{size(): int} & \texttt{add(x)} & $x \in Set$ \\
    & \texttt{remove(x)} & $x \notin Set$ \\
\texttt{contains(x): bool} & \texttt{add(y)} & $x \ne y \lor y \in Set$ \\
    & \texttt{remove(y)} & $x \ne y \lor y \notin Set$ \\
    & \texttt{size()} & $\forall x$ \\
    \hline
\end{tabular}
\caption{\label{spec} Commutativity Specification for Set.}
\end{table}

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in Table \ref{tab:spec}. It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$. In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically.

For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in Table \ref{tab:spec}, but there may be many that express varying degrees of commutativity. The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute. As Kulkarni \cite{Kulkarni:PLDI11} explored, there is a complicated tradeoff between exposing more concurrency and efficiently tracking commutativity. Data structure designers and users navigate these choices to optimize their programs.

## Transactional Boosting
If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record. This technique of raising the level of abstraction in transactions to operations is known as *transactional boosting* \cite{Herlihy:PPoPP08}. This straightforward use of commutativity was shown to significantly improve abort rates in state-of-the-art software transactional memory. In Section \ref{sec:evaluation}, we show how it can be applied to distributed transactions.

>>>>

In situations where consistency really is crucial, developers can enforce stronger guarantees on top of any platform, but they often get it wrong. Google, recognizing the benefit to programmer productivity, reintroduced serializable transactions to their next generation database, Spanner \cite{Spanner}. However, this leaves them with two extreme choices, with significant performance impact. If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model? What if the programmer could express the semantics they desire succinctly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?

@@

Performance benefits come from understanding the properties of ADT operations: those that commute with each other can be performed concurrently, even on multiple copies of the record. This means that transactions whose operations commute abort less, approaching the performance without transactions. This cleanly captures the cases described above where conflicts were unobservable or ordering didn't matter but in a safe way because any operations that don't in fact commute will be handled with traditional concurrency control. Using insights from multi-core concurrent data structures, we show in \autoref{comm} that it is practical to reason about the matrix of commutativity among operations and build implementations that make the right tradeoff between the amount of concurrency allowed and the efficiency of tracking it.

@@

# Commutativity {#comm}

@@

\begin{table}
\resizebox{\columnwidth}{!}{%
\centering
\begin{tabular}{lll}
\textbf{method:} & \textbf{commutes with:} & \textbf{when:} \\
\hline
\texttt{add(x): void} & \texttt{add(y)} & $\forall x, y$ \\
\texttt{remove(x): void} & \texttt{remove(y)} & $\forall x, y$ \\
    & \texttt{add(y)} & $x \ne y$ \\
\texttt{size(): int} & \texttt{add(x)} & $x \in Set$ \\
    & \texttt{remove(x)} & $x \notin Set$ \\
\texttt{contains(x): bool} & \texttt{add(y)} & $x \ne y \lor y \in Set$ \\
    & \texttt{remove(y)} & $x \ne y \lor y \notin Set$ \\
    & \texttt{size()} & $\forall x$ \\
    \hline
\end{tabular}
\caption{\label{spec} Commutativity Specification for Set.}}
\end{table}

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a property of the data type and is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for a data type its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}. It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$. In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically.

For a given data type interface, there is one specification that expresses the maximum concurrency, like the one in \autoref{spec}, but there may be many that express varying degrees of commutativity. The space of possible specifications forms a *lattice* where $\top$ specifies the maximum possible commutativity, and $\bot$ represents the case where no operations are allowed to commute. As Kulkarni \cite{Kulkarni:PLDI11} explored, there is a complicated tradeoff between exposing more concurrency and efficiently tracking commutativity. Data structure designers and users navigate these choices to optimize their programs.

## Transactional Boosting
If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record. This technique of raising the level of abstraction in transactions to operations is known as *transactional boosting* \cite{Herlihy:PPoPP08}. This straightforward use of commutativity was shown to significantly improve abort rates in state-of-the-art software transactional memory. In \autoref{eval}, we show how it can be applied to distributed transactions.
